---
title: "DSC 630"
author: "Adam Curry"
date: "11/18/2020"
output: word_document
---

## Import the libraries and the data from SQL
```{r setup, include=FALSE}
#import libraries
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
# create the connection to the database
db <- dbConnect( SQLite(), dbname = "C:\\Users\\adamp\\df_Joiner.db" )

# localize the data
dbListTables(db)
df <- dbReadTable(db, "df_to_R")
```
## Apply the algorithms
k-means cluster variables based on economic parameters, stringency index, and google retail and recreation 
```{r}
set.seed(123)
df_K <- select(df,'econ_derived','StringencyIndexForDisplay','gps_retail_and_recreation')
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
df_K <- na.omit(df_K)
# convert all the columns to numeric
df_K <- sapply(df_K, as.numeric)
df_K <- scale(df_K)
df_K <- data.frame(df_K)
```

using the average silhoutte method method, find the optimal number of clusters for the K-means cluster algorithm.
create the K-Means algorithm with 3 centroids. The first step when using k-means clustering is to indicate the number of clusters (k) that will be generated in the final solution.
```{r}
fviz_nbclust(df_K, kmeans, method = "silhouette")


k2 <- kmeans(df_K, centers = 3, nstart = 25)

```

# plot the clusters   
fviz_cluster will perform principal component analysis (PCA) and plot the data points according to the first two principal components that explain the majority of the variance.

```{r}
fviz_cluster(k2, geom = "point", data = df_K) + 
  ggtitle("K-Means cluster of observations utilizing economic metrics, 
          government stringency, and workplace mobility")
#the number of clusters (k) that will be generated in the final solution.

```

# add the clusters to the data frame
```{r}
df$cluster_economy <- k2$cluster

# split apart clusters into separate columns

# replace the NA's in the new columns with 0's
df[is.na(df)] <- 0
df$cluster <- k2$cluster

# write the clustered data back to database to continue work in Python
dbRemoveTable(db, "db_clustered", df)

dbWriteTable(db, "db_clustered", df)

# Disconnect from the database
dbDisconnect(db)
```

